---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk[['set']](
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# `gips`

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version-last-release/gips)](https://CRAN.R-project.org/package=gips)
[![R-CMD-check](https://github.com/PrzeChoj/gips/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PrzeChoj/gips/actions/workflows/R-CMD-check.yaml)
[![test-coverage](https://codecov.io/gh/PrzeChoj/gips/branch/main/graph/badge.svg)](https://codecov.io/gh/PrzeChoj/gips?branch=main)
<!-- badges: end -->

gips - Gaussian model Invariant by Permutation Symmetry

`gips` is an R package that finds the permutation symmetry group such that the covariance matrix of the given data is invariant under it. Knowledge of such a permutation can drastically decrease the number of parameters needed to fit the model. That means that with `gips`, it is possible to find the Gaussian model with more parameters than the number of observations. Sometimes, even if the number of observations is bigger than the number of parameters, the covariance matrix found with `gips` better approximates the actual covariance behind the data.

## `gips` will help you with two things:
1. Exploratory Data Analysis - with `gips`, you can find the permutation of features that approximately does not change the covariance matrix.
2. Modeling - with `gips`, you can accurately use the found permutation to fit the normal models like LDA or QDA.

## Installation

You can install the development version of gips from [GitHub](https://github.com/PrzeChoj/gips) with:

``` r
# install.packages("devtools")
devtools::install_github("PrzeChoj/gips")
```

## Example

TODO(Example without assuming known mean)

Assume we know the mean is $0$, and we want to estimate the covariance matrix:

```{r example}
library(gips)

# Prepare model, multivariate normal distribution
perm_size <- 6
mu <- numeric(perm_size)  
sigma_matrix <- matrix(
  data = c(
    1.0, 0.8, 0.6, 0.4, 0.6, 0.8,
    0.8, 1.0, 0.8, 0.6, 0.4, 0.6,
    0.6, 0.8, 1.0, 0.8, 0.6, 0.4,
    0.4, 0.6, 0.8, 1.0, 0.8, 0.6,
    0.6, 0.4, 0.6, 0.8, 1.0, 0.8,
    0.8, 0.6, 0.4, 0.6, 0.8, 1.0
  ),
  nrow = perm_size, byrow = TRUE
) # sigma_matrix is a matrix invariant under permutation (1,2,3,4,5,6)
number_of_observations <- 13

# generate example data from a model:
Z <- MASS::mvrnorm(number_of_observations, mu = mu, Sigma = sigma_matrix)

# calculate the covariance matrix from the data:
S <- (t(Z) %*% Z) / number_of_observations
# the theoretical mean is 0, so this S is the Maximum likelihood estimation of the covariance matrix

# Make the gips object out of data:
g <- gips(S, number_of_observations, was_mean_estimated = FALSE)

# Find the Maximum A Posteriori Estimator for the permutation:
g_map <- find_MAP(g, show_progress_bar = TRUE, optimizer = "full")
g_map

summary(g_map)

# find_MAP() found the permutation; now, make a matrix estimator:
project_matrix(S, g_map[[1]])

# Plot the found matrix:
plot(g_map, type = "heatmap")
```

# Credits

It was developed in 2022 by Przemysław Chojecki and Paweł Morgen under the leadership of Ph.D. Bartosz Kołodziejek within the "CyberiADa-3" (2021) grant from the Warsaw University of Technology.

