---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk[["set"]](
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  cache = TRUE
)
```

# `gips` <a href="https://przechoj.github.io/gips/"><img src="man/figures/logo.svg" align="right" height="139" /></a>

<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![CRAN status](https://www.r-pkg.org/badges/version/gips)](https://CRAN.R-project.org/package=gips)
[![R-CMD-check](https://github.com/PrzeChoj/gips/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PrzeChoj/gips/actions/workflows/R-CMD-check.yaml)
[![test-coverage](https://codecov.io/gh/PrzeChoj/gips/branch/main/graph/badge.svg)](https://app.codecov.io/gh/PrzeChoj/gips?branch=main)

<!-- badges: end -->

gips - Gaussian model Invariant by Permutation Symmetry

`gips` is an R package that looks for permutation symmetries in the multivariate Gaussian sample. Such symmetries reduce the free parameters in the unknown covariance matrix. This is especially useful when the number of variables is substantially larger than the number of observations.


## `gips` will help you with two things:
1. Finding hidden symmetries between the variables. `gips` can be used as an exploratory tool for searching the space of permutation symmetries of the Gaussian vector.  Useful in the Exploratory Data Analysis (EDA).
2. Covariance estimation. The MLE for the covariance matrix is known to exist if and only if the number of variables is less or equal to the number of observations. Additional knowledge of symmetries significantly weakens this requirement. Moreover, the reduction of model dimension brings the advantage in terms of precision of covariance estimation. 

## Installation

From [CRAN](https://CRAN.R-project.org/package=gips):
``` r
# Install the released version from CRAN:
install.packages("gips")
```

From [GitHub](https://github.com/PrzeChoj/gips):
``` r
# Install the development version from GitHub:
# install.packages("devtools")
devtools::install_github("PrzeChoj/gips")
```

## Examples

### Example 1 - EDA

Assume we have the data, and we want to understand its structure:

```{r example_mean_unknown0}
library(gips)

Z <- HSAUR::aspirin

# Renumber the columns for better readability:
Z[,c(2,3)] <- Z[,c(3,2)]
```

```{r example_mean_unknown1, include=FALSE, dependson="example_mean_unknown0"}
names(Z) <- NULL
my_add_text <- function(gg_plot_object){
  my_col_names <- c("deaths after placebo", "deaths after Aspirin",
                    "treated with placebo", "treated with Aspirin")
  
  suppressMessages( # message from ggplot2
    out <- gg_plot_object +
      ggplot2::scale_x_continuous(labels = my_col_names,
                                  breaks = 1:4) +
      ggplot2::scale_y_reverse(labels = my_col_names,
                               breaks = 1:4) +
      ggplot2::theme(axis.text.x = ggplot2::element_text(angle=8))
  )
  
  out
}
```

Assume the data `Z` is normally distributed.

```{r example_mean_unknown2, dependson="example_mean_unknown1"}
dim(Z)
number_of_observations <- nrow(Z) # 7
perm_size <- ncol(Z) # 4

S <- cov(Z)
round(S)

g <- gips(S, number_of_observations)
my_add_text(plot(g, type = "heatmap"))
```

Remember, we analyze the covariance matrix. We can see some strong similarities between the covariances of columns 3 and 4. Those have similar variances (`S[3,3]` $\approx$ `S[4,4]`), and their covariances with the rest of the columns are alike (`S[1,3]` $\approx$ `S[1,4]` and `S[2,3]` $\approx$ `S[2,4]`).

Note that the variances of columns 1 and 2 are also similar, but the covariances with other columns (3 and 4) are not alike.

Let's see if the `find_MAP()` will find this relationship:

```{r example_mean_unknown3, dependson="example_mean_unknown2"}
g_MAP <- find_MAP(g, optimizer = "brute_force")

g_MAP
```

The `find_MAP` found the relationship (3,4). In its opinion, the variances [3,3] and [4,4] are so close to each other that it is reasonable to consider them equal. Similarly, the covariances [1,3] and [1,4]; just as covariances [2,3] and [3,4], also will be considered equal:

```{r example_mean_unknown4, dependson="example_mean_unknown3"}
S_projected <- project_matrix(S, g_MAP[[1]])
round(S_projected)

my_add_text(plot(g_MAP, type = "heatmap"))
```

This `S_projected` matrix can now be interpreted as a more stable covariance matrix estimator.

We can also interpret the data suggesting there is, for example, the same covariance of "number of deaths after Aspirin" with "number of people treated with \*" no matter if the "\*" represents the placebo or Aspirin.

### Example 2 - modeling

First, construct data for the example:
```{r example_mean_known1}
# Prepare model, multivariate normal distribution
perm_size <- 6
mu <- numeric(perm_size)  
sigma_matrix <- matrix(
  data = c(
    1.1, 0.8, 0.6, 0.4, 0.6, 0.8,
    0.8, 1.1, 0.8, 0.6, 0.4, 0.6,
    0.6, 0.8, 1.1, 0.8, 0.6, 0.4,
    0.4, 0.6, 0.8, 1.1, 0.8, 0.6,
    0.6, 0.4, 0.6, 0.8, 1.1, 0.8,
    0.8, 0.6, 0.4, 0.6, 0.8, 1.1
  ),
  nrow = perm_size, byrow = TRUE
) # sigma_matrix is a matrix invariant under permutation (1,2,3,4,5,6)


# Generate example data from a model:
Z <- MASS::mvrnorm(4, mu = mu, Sigma = sigma_matrix)
# End of prepare model
```

Suppose we do not know the true covariance matrix $\Sigma$ and we want to estimate it. We cannot use the standard MLE because it does not exists ($4 < 6$, $n < p$).

We will assume it was generated from the normal distribution with the mean $0$.
```{r example_mean_known2, dependson="example_mean_known1"}
library(gips)
dim(Z)
number_of_observations <- nrow(Z) # 4
perm_size <- ncol(Z) # 6

# Calculate the covariance matrix from the data:
S <- (t(Z) %*% Z) / number_of_observations
```

Make the gips object out of data:
```{r example_mean_known3, dependson="example_mean_known2"}
g <- gips(S, number_of_observations, was_mean_estimated = FALSE)
```

Find the Maximum A Posteriori Estimator for the permutation. Space is small ($6! = 720$), so it is reasonable to browse the whole of it:
```{r example_mean_known4, dependson="example_mean_known3"}
g_map <- find_MAP(g, optimizer = "brute_force")
g_map
```

We see that the found permutation is hundreds of times more likely than making no additional assumption. That means the additional assumptions are justified.
```{r example_mean_known5, dependson="example_mean_known4"}
summary(g_map)$n0
summary(g_map)$n0 <= number_of_observations # 1 <= 4
```

What is more, we see the number of observations ($4$) is bigger or equal to $n_0 = 1$, so we can estimate the covariance matrix with the Maximum Likelihood estimator:
```{r example_mean_known6, dependson="example_mean_known5"}
S_projected <- project_matrix(S, g_map[[1]])
S_projected

# Plot the found matrix:
plot(g_map, type = "heatmap")
```

We see `gips` found the data's structure, and we could estimate the covariance matrix with huge accuracy only with a small amount of data and additional reasonable assumptions.

Note that the rank of the `S` matrix was 4, while the rank of the `S_projected` matrix was 6.


# Further reading

For more examples and introduction, see `vignette("gips", package="gips")` or its [pkgdown page](https://przechoj.github.io/gips/articles/gips.html).


# Acknowledgment

Project was funded by Warsaw University of Technology within the Excellence Initiative: Research University (IDUB) programme.

