---
title: "gips"
author: "Przemysław Chojecki, Paweł Morgen, Bartosz Kołodziejek"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gips}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk[["set"]](
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)

options(scipen=999) # turn off scientific notation
```

# The problem

Quite often, we have too little data to perform valid inference. Consider the 
situation with multivariate Gaussian distribution, where we have few observations 
compared to the number of variables. That's the case for 
graphical models used in biology or medicine. In such a setting, the usual way
of finding the covariance matrix (the maximum likelihood method) isn't statistically 
applicable. What now?

# Invariance by permutation

In some cases, the interchange of variables in the vector does not change its distribution.
In the multivariate Gaussian case, it would mean that they have the same variances and covariances with other respective variables. For instance, in the following covariance matrix, variables X1 and X3 are interchangeable, meaning that vectors (X1, X2, X3) and (X3, X2, X1) have the same distribution.

```{r symvariant_matrix, echo=FALSE}
X <- matrix(c(1, 2, 3,
         2, 4, 2,
         3, 2, 1), byrow=TRUE, ncol=3, dimnames=list(c('X1', 'X2', 'X3'), 
                                                     c('X1', 'X2', 'X3')))
heatmap(X, Rowv = NA, Colv = NA, main = "", symm=TRUE)
```

Now, we can state this interchangeability property in terms of permutations. In our case, the distribution of (X1, X2, X3) is **invariant by permutation** (1->3, 3->1), or in cyclic form (1,3). This is equivalent to saying that swapping the first with the third row and then swapping the first and third columns of covariance matrix results in the same matrix. Then we say that this covariance matrix is **invariant by permutation**.

Of course, in the samples collected in the real world, no perfect equalities will be observed. 
Still, if the respective values in the (poorly) estimated covariance matrix were close, adopting
a particular assumption about invariance by permutation would be a reasonable step.

# Package `gips`

We propose creating a set of constraints on the covariance matrix so that the 
maximum likelihood method can be used after all. The constraint we consider is - none other than - 
invariance under permutation symmetry. 

This package provides a way to find a *reasonable* permutation to be used as a constraint in covariance matrix estimation. In this case, *reasonable* means such that maximizes the Bayesian posterior distribution when using a Wishart-like distribution on symmetric, positive definite matrices as a prior. The idea, exact formulas, and algorithm sketch contribute to PAPER.

# Example

<div id="spoiler1" style="display:none">
```{r toy_example_data_making, include = TRUE}
perm_size <- 4
mu <- runif(4, -10, 10) # Assume we don't know the mean
sigma_matrix <- matrix(c(6.25, 2,   0.5, 0.5,
                         2,    3,   0.3, 0.3,
                         0.5,  0.3, 1,   0.8,
                         0.5,  0.3, 0.8, 1), ncol=4)
# sigma_matrix is a matrix invariant under permutation (3,4)
number_of_observations <- 9
toy_example_data <- withr::with_seed(1234,
    code = MASS::mvrnorm(number_of_observations,
                         mu = mu, Sigma = sigma_matrix)
)
```
</div>

<button title="Click to show the data preparation" type="button"
   onclick="if(document.getElementById('spoiler1') .style.display=='none')
              {document.getElementById('spoiler1') .style.display=''}
            else{document.getElementById('spoiler1') .style.display='none'}">
  Show/hide data preparation
</button>

```{r toy_example_data_show, fig_width = 0.3, fig.align = "center"}
library(gips)

toy_example_data
S <- cov(toy_example_data)

g <- gips(S = S, number_of_observations = nrow(toy_example_data))

plot(g, type = "heatmap")
g_map <- find_MAP(g, optimizer = "brute_force", return_probabilities = TRUE, save_all_perms = TRUE)

plot(g_map, type = "heatmap")
get_probabilities_from_gips(g_map)
```


# Further reading

1. To learn more about the available optimizers in `find_MAP()` and how to use those, see `vignette("Optimizers")` or its [pkgdown page](https://przechoj.github.io/gips/articles/Optimizers.html).
2. To learn more about the math and theory behind the `gips` package, see `vignette("Theory")` or its [pkgdown page](https://przechoj.github.io/gips/articles/Theory.html).
