---
title: "Introduction"
author: "Paweł Morgen, Przemysław Chojecki"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## The problem

Quite often, we have too little data to perform valid inference. Consider the 
situation with multivariate Gaussian distribution, where we have few observations 
compared to the number of variables. That's the case for 
graphical models used in biology or medicine. In such a setting, the usual way
of finding the covariance matrix (the maximum likelihood method) isn't statistically 
applicable. What now?

## Invariance by permutation

Quite often, some variables can be used interchangeably. In the multivariate Gaussian 
case, it would mean that they have the same variances and covariances with other respective variables. For instance, in the following matrix, variables X1 and X3 are interchangeable:

```{r symvariant_matrix, echo=FALSE}
X <- matrix(c(1, 2, 3,
         2, 4, 2,
         3, 2, 1), byrow=TRUE, ncol=3, dimnames=list(c('X1', 'X2', 'X3'), 
                                                     c('X1', 'X2', 'X3')))
heatmap(X, Rowv = NA, Colv = NA, main = "", symm=TRUE)
```

Now, we can state this interchangeability property in terms of permutations. In our case,
swapping the first with the third row and then swapping the first and third columns results
in the same matrix. In other words, this matrix is **invariant by permutation** (1->3, 3->1). Or in cyclic form (1,3).

Of course, in the samples collected in the real world, no perfect equalities will be observed. 
Still, if the respective values in the (poorly) estimated covariance matrix were close, adopting
a particular assumption about invariance by permutation would be a reasonable step.

## Package `gips`

We propose creating a set of constraints on the covariance matrix so that the 
maximum likelihood method can be used after all. The constraint we consider is - none other than - 
invariance under permutation symmetry. 

This package provides a way to find a *reasonable* permutation to be used as a constraint in covariance matrix estimation. In this case, *reasonable* means such that maximizes the Bayesian posterior distribution when using Wishart distribution on symmetric, positive definite matrices as a prior. The idea, exact formulas, and algorithm sketch contribute to PAPER.

## Example


```{r setup}
library(gips)
```
